---
title: "raincloudplots"
author: "Jordy van Langen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{raincloudplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(raincloudplots)
library(ggplot2)
```


This vignette demonstrates the different types of *raincloudplots* that one can make with this package.

## Background

We have developed this package to visualize data in rainclouds, which are, for instance, suited for repeated measures data. 

- It all started with the [pre-print of raincloudplots](https://wellcomeopenresearch.org/articles/4-63) in 2019 and accompanying [GitHub repository RainCloudPlots](https://github.com/RainCloudPlots/RainCloudPlots)
- In the beginning of 2020, a tutorial called [‘open-visualizations’](https://github.com/jorvlan/open-visualizations) was released and turned out to be a valuable addition to the previously published pre-print. This tutorial provides detailed and extensive R code to create robust and transparent repeated measures visualizations, by showing the slope change for each individual data point over time. To date (30-11-2020), this tutorial has been cited in 15 scientific papers. However, using this tutorial requires sufficient R programming knowledge and might therefore not be suitable for non-R experts. 
- Therefore, we have created this dedicated [raincloudplots](https://github.com/jorvlan/raincloudplots) package. This package is tailored towards easy visualization of grouped and repeated measures data. Moreover, it also provides individually linked repeated measures visualizations, which add detail and richness to a multitude of between/within-subject designs. Researchers that would like to visualize more complex repeated measures designs, for instance with more groups and more time-points, please see our extensive tutorials:
  - [https://github.com/RainCloudPlots/RainCloudPlots](https://github.com/RainCloudPlots/RainCloudPlots)
  - [https://github.com/jorvlan/open-visualizations](https://github.com/jorvlan/open-visualizations)

## data_1x1

- `data_1x1` creates the long-format data.frame that is needed for 1-by-1 rainclouds.
  - `array_1` the first array of datapoints to be plotted
  - `array_2` the second array of datapoints to be plotted
  - `jit_distance` the amount of distance between jittered datapoints (0.9 by default)
  - `jit_seed` the amount used in set.seed() (321 by default)

```{r}
df_1x1 <- data_1x1(
  array_1 = iris$Sepal.Length[1:50],
  array_2 = iris$Sepal.Length[51:100],
  jit_distance = .09,
  jit_seed = 321)

head(df_1x1)
tail(df_1x1)
```

## raincloud_1x1

- `raincloud_1x1` creates the 1-by-1 comparison for grouped data.
  - `data` the data.frame created with `data_1x1`
  - `colors` concatenated string for both colors
  - `fills` concatenated string for both both fills
  - `size` size of the data 
  - `alpha` alpha of the data
  - `ort` vertical or horizontal display of rainclouds

```{r}
raincloud_1_h <- raincloud_1x1(
  data = df_1x1, 
  colors = (c('dodgerblue','darkorange')), 
  fills = (c('dodgerblue','darkorange')), 
  size = 1, 
  alpha = .6, 
  ort = 'h') +

scale_x_continuous(breaks=c(1,2), labels=c("Group1", "Group2"), limits=c(0, 3)) +
  xlab("Groups") + 
  ylab("Score") +
  theme_classic()

raincloud_1_h
```

```{r}
raincloud_1_v <- raincloud_1x1(
  data = df_1x1, 
  colors = (c('dodgerblue','darkorange')), 
  fills = (c('dodgerblue','darkorange')), 
  size = 1, 
  alpha = .6, 
  ort = 'v') +

scale_x_continuous(breaks=c(1,2), labels=c("Group1", "Group2"), limits=c(0, 3)) +
  xlab("Groups") + 
  ylab("Score") +
  theme_classic()

raincloud_1_v
```

## raincloud_1x1_repmes

- `raincloud_1x1_repmes` creates a 1-by-1 repeated measures raincloud.
  - `data` the data.frame created with `data_1x1`
  - `colors` concatenated string for both colors
  - `fills`  concatenated string for both fills
  - `line_color` color of the lines
  - `line_alpha` alpha of the lines
  - `size` size of the data
  - `alpha` alpha of the data
  - `align_clouds` FALSE if spreaded on different x-axis ticks, TRUE if aligned on same x-axis tick

```{r}
raincloud_2 <- raincloud_1x1_repmes(
  data = df_1x1,
  colors = (c('dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange')),
  line_color = 'gray',
  line_alpha = .3,
  size = 1,
  alpha = .6,
  align_clouds = FALSE) +
 
scale_x_continuous(breaks=c(1,2), labels=c("Pre", "Post"), limits=c(0, 3)) +
  xlab("Time") + 
  ylab("Score") +
  theme_classic()

raincloud_2
```

```{r}
raincloud_2_aligned <- raincloud_1x1_repmes(
  data = df_1x1,
  colors = (c('dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange')),
  line_color = 'gray',
  line_alpha = .3,
  size = 1,
  alpha = .6,
  align_clouds = TRUE) +
 
scale_x_continuous(breaks=c(1,2), labels=c("Pre", "Post"), limits=c(0, 3)) +
  xlab("Time") + 
  ylab("Score") +
  theme_classic()

raincloud_2_aligned
```

## data_2x2

- `data_2x2` creates the long-format data.frame needed for 2 x 2 repeated measures rainclouds.
  - `array_1` the first array of datapoints to be plotted
  - `array_2` the second array of datapoints to be plotted
  - `array_3` the array of datapoints to be plotted
  - `array_4` the array of datapoints to be plotted
  - `array_5` the array of datapoints to be plotted (OPTIONAL: only needed for 2x3 repeated measures, see below)
  - `array_6` the array of datapoints to be plotted (OPTIONAL: only needed for 2x3 repeated measures, see below)
  - `labels` concatenated string of both group labels
  - `spread_x_ticks` FALSE if 2 x-ticks, TRUE if 4 x-ticks
  - `jit_distance` the amount of distance between jittered datapoints (0 by default)
  - `jit_seed` the amount used in set.seed() (321 by default)
  
```{r}
df_2x2 <- data_2x2(
  array_1 = iris$Sepal.Length[1:50],
  array_2 = iris$Sepal.Length[51:100],
  array_3 = iris$Sepal.Length[101:150],
  array_4 = iris$Sepal.Length[81:130],
  labels = (c('congruent','incongruent')),
  jit_distance = .09,
  jit_seed = 321,
  spread_x_ticks = FALSE) 

head(df_2x2)
tail(df_2x2)
```

```{r}
df_2x2_spread <- data_2x2(
  array_1 = iris$Sepal.Length[1:50],
  array_2 = iris$Sepal.Length[51:100],
  array_3 = iris$Sepal.Length[101:150],
  array_4 = iris$Sepal.Length[81:130],
  labels = (c('congruent','incongruent')),
  jit_distance = .09,
  jit_seed = 321,
  spread_x_ticks = TRUE) 

head(df_2x2_spread)
tail(df_2x2_spread)
```

## raincloud_2x2_repmes

- `raincloud_2x2_repmes` creates the 2 x 2 repeated measures raincloud.
  - `data` the data.frame created with `data_2x2`
  - `colors` concatenated string for both colors
  - `fills`  concatenated string for both fills
  - `line_color` color of the lines (Only when `spread_x_ticks` TRUE)
  - `line_alpha` alpha of the lines (Only when `spread_x_ticks` TRUE)
  - `size` size of the data
  - `alpha` alpha of the data
  - `spread_x_ticks` FALSE if 2 x-ticks, TRUE if 4 x-ticks
  
```{r}
raincloud_2x2 <- raincloud_2x2_repmes(
  data = df_2x2,
  colors = (c('dodgerblue', 'darkorange', 'dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange', 'dodgerblue', 'darkorange')),
  size = 1,
  alpha = .6,
  spread_x_ticks = FALSE) +

scale_x_continuous(breaks=c(1,2), labels=c("Pre", "Post"), limits=c(0, 3)) +
  xlab("Time") + 
  ylab("Score") +
  theme_classic()

raincloud_2x2

```

```{r}
raincloud_2x2_spread <- raincloud_2x2_repmes(
  data = df_2x2_spread,
  colors = (c('dodgerblue', 'darkorange', 'dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange', 'dodgerblue', 'darkorange')),
  line_color = 'gray',
  line_alpha = .3,
  size = 1,
  alpha = .6,
  spread_x_ticks = TRUE) +

scale_x_continuous(breaks=c(1,2,3,4), labels=c("Pre", "Post", "Pre", "Post"), limits=c(0, 5)) +
  xlab("Time") + 
  ylab("Score") +
  theme_classic()

raincloud_2x2_spread
```

## raincloud_2x3_repmes

```{r}
df_2x3 <- data_2x2(
  array_1 = iris$Sepal.Length[1:50],
  array_2 = iris$Sepal.Length[51:100],
  array_3 = iris$Sepal.Length[101:150],
  array_4 = iris$Sepal.Length[81:130],
  array_5 = iris$Sepal.Length[21:70],
  array_6 = iris$Sepal.Length[41:90],
  labels = (c('congruent','incongruent')),
  jit_distance = .05,
  jit_seed = 321) 

head(df_2x3)
tail(df_2x3)
```



```{r}
raincloud_2x3_vertical <- raincloud_2x3_repmes(
  data = df_2x3,
  colors = (c('dodgerblue', 'darkorange', 'dodgerblue',
              'darkorange', 'dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange', 'dodgerblue',
             'darkorange', 'dodgerblue', 'darkorange')),
  size = 1,
  alpha = .6,
  ort = 'v') +

scale_x_continuous(breaks=c(1,2,3), labels=c("T-1", "T-2", "T-3"), limits=c(0, 4)) +
  xlab("Time") + 
  ylab("Score") +
  theme_classic()

raincloud_2x3_vertical
```

```{r}
raincloud_2x3_horizontal <- raincloud_2x3_repmes(
  data = df_2x3,
  colors = (c('dodgerblue', 'darkorange', 'dodgerblue',
              'darkorange', 'dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange', 'dodgerblue',
             'darkorange', 'dodgerblue', 'darkorange')),
  size = 1,
  alpha = .6,
  ort = 'h') +

scale_x_continuous(breaks=c(1,2,3), labels=c("T-1", "T-2", "T-3"), limits=c(0, 4)) +
  xlab("Time") + 
  ylab("Score") +
  theme_classic()

raincloud_2x3_horizontal
```

